\textbf{SPECFEM3D}\newline \textbf{Cartesian}\newline \textbf{User Manual}

### 2022-11-12 doc/USER_MANUAL/05_running_the_solver.tex
updates tex files (\\ to \newline) for better pandoc conversions to wiki pages

Now that you have successfully generated the databases, you are ready
to compile the solver. In the main directory, type
{\small
\begin{verbatim}
  make xspecfem3D
\end{verbatim}
}
\noindent
Please note that \texttt{xspecfem3D} must be called directly from
the main directory, as most of the binaries of the package.\\

@@

\noindent
Most parameters in the \texttt{DATA/Par\_file} should be set prior to running
the databases generation. Only the following parameters may be changed
after running \texttt{xgenerate\_databases}:
\begin{itemize}
\item the simulation type control parameters \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the absorbing boundary control parameter \texttt{PML\_CONDITIONS}
on condition that the\\
 \texttt{PML\_INSTEAD\_OF\_FREE\_SURFACE} flag remains unmodified
after running the databases generation.
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered{} option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{MOVIE\_TYPE}, \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO} and\\
\texttt{NTSTEP\_BETWEEN\_OUTPUT\_SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{DATA/Par\_file} implies rerunning both
the database generator \texttt{xgenerate\_databases} and the solver
\texttt{xspecfem3D}.\\

@@

\noindent
Each line represents one station in the following format:
{\small
\begin{verbatim}
Station Network Latitude(degrees) Longitude(degrees) Elevation(m) burial(m)
\end{verbatim}
}
The solver \texttt{xspecfem3D} filters the list of stations in file
\texttt{DATA/STATIONS} to exclude stations that are not located within
the region given in the \texttt{Mesh\_Par\_file} (between \texttt{LATITUDE\_MIN}
and \texttt{LATITUDE\_MAX} and between \texttt{LONGITUDE\_MIN} and
\texttt{LONGITUDE\_MAX}). The filtered file is called \texttt{DATA/STATIONS\_FILTERED}.
Elevation and burial are generally applicable to geographical regions.
Burial is measured down from the top surface.\\

\noindent
For other problems in other
fields (ultrasonic testing, medical imaging etc...),
it may be confusing. We generally follow either one of the following
procedures for those kind of problems:
\begin{itemize}
\item {\it Procedure 1:} mostly for geophysics, when the top surface is a free surface (topography) and the five other edges of the mesh are absorbing surfaces
\begin{enumerate}
\item [-] Put the origin on the top of the model.\\
\item [-] Let's say you want to place two receivers at (x1,y1,z1) and
(x2,y2,z2). Your STATIONS file should look like:
{\small \begin{verbatim}
BONE  GR  y1  x1  0.00  -z1
BONE  GR  y2  x2  0.00  -z2
\end{verbatim}}
\end{enumerate}

\item {\it Procedure 2:} useful for other application domains, in which using the absolute $Z$ position of the sources and receivers is more standard than using their depth from the surface
\begin{enumerate}
\item [-] In principle in the international CMTSOLUTION format in geophysics
the depth is given in kilometers; however for users in other fields (non-destructive testing, medical imaging, near-surface studies...) who may prefer to give the position of the source (rather than its depth from the surface), or for people who use FORCESOLUTION to describe the source rather than CMTSOLUTION, we provide an option called USE\_SOURCES\_RECEIVERS\_Z in the \texttt{DATA/Par\_file}, and if so that position is read from CMTSOLUTION in meters rather than kilometers (and again, it is then the true position in the mesh, not the depth). When option USE\_SOURCES\_RECEIVERS\_Z in the \texttt{DATA/Par\_file} is on, this remark applies to the position of the receivers as well.\\
\item [-] Let's say you want to place two receivers at (x1,y1,z1) and
(x2,y2,z2). Your STATIONS file should then look like:
{\small \begin{verbatim}
BONE  GR  y1  x1  0.00  z1
BONE  GR  y2  x2  0.00  z2
\end{verbatim}}
The option USE\_SOURCES\_RECEIVERS\_Z set to .true. will then discard the elevation and set
burial as the $z$ coordinate.
Third column is Y and Fourth is X due to the latitude/longitude convention.\\
\end{enumerate}

You can replace the station name "BONE" with any word of length less
than 32, and the network name "GR" with any word of length less than
8.
And you can always plot OUTPUT\_FILES/sr.vtk file in ParaView to check the
source/receiver locations after your simulation.\\
\end{itemize}

@@

\begin{itemize}
\item The simulation is launched within the root directory \texttt{EXAMPLE\_ROOT\_DIR}\\
(usually \texttt{mpirun -np N ./bin/xspecfem3D}).
\item \texttt{DATA} should contain the \texttt{Par\_file} parameter file with
\texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS} as explained in Chapter~\ref{cha:Creating-Distributed-Databases}.
\item \texttt{DATABASES\_MPI} and {OUTPUT\_FILES} directory may contain the mesher output but they
are not required as they are superseded by the ones in the \texttt{runXXX} directories.
\item \texttt{runXXXX} directories must be created beforehand. There should be be as many as
\texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS} and the numbering should be contiguous, starting from \texttt{0001}.
They all should have \texttt{DATA},
\texttt{DATABASES\_MPI} and \texttt{OUTPUT\_FILES} directories. Additionally a \texttt{SEM} directory
containing adjoint sources have to be created to perform adjoint simulations.
\item \texttt{runXXXX/DATA} directories must all contain a \texttt{CMTSOLUTION} file,
a \texttt{STATIONS} file along with an eventual \texttt{STATIONS\_ADJOINT} file.
\item If \texttt{BROADCAST\_SAME\_MESH\_AND\_MODEL} is set to \texttt{.true.} in \texttt{DATA/Par\_file},
only \texttt{run0001/OUTPUT\_FILES} and \texttt{run0001/DATABASES\_MPI} directories need to contain the files
outputted by the mesher.
\item If \texttt{BROADCAST\_SAME\_MESH\_AND\_MODEL} is set to \texttt{.false.} in \texttt{DATA/Par\_file},
every \texttt{runXXXX/OUTPUT\_FILES} and \texttt{runXXXX/DATABASES\_MPI} directories need to contain the files
outputted by the mesher. Note that while the meshes might have been created from different models and
parameter sets, they should have been created using the same number of MPI processes.
\end{itemize}

@@

\noindent
The model used is a constant $Q$, thus with no dependence on frequency ($Q(f)$ = constant).
See e.g. \cite{BlKoChLoXi16}. \\

\noindent
However in practice for technical reasons it is approximated based on the sum of different Generalized Zener body mechanisms
and thus the code outputs the band in which the approximation is very good, outside of that range it can be less accurate.
The logarithmic center of that frequency band is the \texttt{ATTENUATION\_f0} parameter defined (in Hz) in input file \texttt{DATA/Par\_file}. \\

>>>>

Now that you have successfully generated the databases, you are ready
to compile the solver. In the main directory, type
{\small
\begin{verbatim}
  make xspecfem3D
\end{verbatim}
}
\noindent
Please note that \texttt{xspecfem3D} must be called directly from
the main directory, as most of the binaries of the package.\newline

@@

\noindent
Most parameters in the \texttt{DATA/Par\_file} should be set prior to running
the databases generation. Only the following parameters may be changed
after running \texttt{xgenerate\_databases}:
\begin{itemize}
\item the simulation type control parameters \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the absorbing boundary control parameter \texttt{PML\_CONDITIONS}
on condition that the\newline
 \texttt{PML\_INSTEAD\_OF\_FREE\_SURFACE} flag remains unmodified
after running the databases generation.
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered{} option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{MOVIE\_TYPE}, \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO} and\newline
\texttt{NTSTEP\_BETWEEN\_OUTPUT\_SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{DATA/Par\_file} implies rerunning both
the database generator \texttt{xgenerate\_databases} and the solver
\texttt{xspecfem3D}.\newline

@@

\noindent
Each line represents one station in the following format:
{\small
\begin{verbatim}
Station Network Latitude(degrees) Longitude(degrees) Elevation(m) burial(m)
\end{verbatim}
}
The solver \texttt{xspecfem3D} filters the list of stations in file
\texttt{DATA/STATIONS} to exclude stations that are not located within
the region given in the \texttt{Mesh\_Par\_file} (between \texttt{LATITUDE\_MIN}
and \texttt{LATITUDE\_MAX} and between \texttt{LONGITUDE\_MIN} and
\texttt{LONGITUDE\_MAX}). The filtered file is called \texttt{DATA/STATIONS\_FILTERED}.
Elevation and burial are generally applicable to geographical regions.
Burial is measured down from the top surface.\newline

\noindent
For other problems in other
fields (ultrasonic testing, medical imaging etc...),
it may be confusing. We generally follow either one of the following
procedures for those kind of problems:
\begin{itemize}
\item {\it Procedure 1:} mostly for geophysics, when the top surface is a free surface (topography) and the five other edges of the mesh are absorbing surfaces
\begin{enumerate}
\item [-] Put the origin on the top of the model.\newline
\item [-] Let's say you want to place two receivers at (x1,y1,z1) and
(x2,y2,z2). Your STATIONS file should look like:
{\small \begin{verbatim}
BONE  GR  y1  x1  0.00  -z1
BONE  GR  y2  x2  0.00  -z2
\end{verbatim}}
\end{enumerate}

\item {\it Procedure 2:} useful for other application domains, in which using the absolute $Z$ position of the sources and receivers is more standard than using their depth from the surface
\begin{enumerate}
\item [-] In principle in the international CMTSOLUTION format in geophysics
the depth is given in kilometers; however for users in other fields (non-destructive testing, medical imaging, near-surface studies...) who may prefer to give the position of the source (rather than its depth from the surface), or for people who use FORCESOLUTION to describe the source rather than CMTSOLUTION, we provide an option called USE\_SOURCES\_RECEIVERS\_Z in the \texttt{DATA/Par\_file}, and if so that position is read from CMTSOLUTION in meters rather than kilometers (and again, it is then the true position in the mesh, not the depth). When option USE\_SOURCES\_RECEIVERS\_Z in the \texttt{DATA/Par\_file} is on, this remark applies to the position of the receivers as well.\newline
\item [-] Let's say you want to place two receivers at (x1,y1,z1) and
(x2,y2,z2). Your STATIONS file should then look like:
{\small \begin{verbatim}
BONE  GR  y1  x1  0.00  z1
BONE  GR  y2  x2  0.00  z2
\end{verbatim}}
The option USE\_SOURCES\_RECEIVERS\_Z set to .true. will then discard the elevation and set
burial as the $z$ coordinate.
Third column is Y and Fourth is X due to the latitude/longitude convention.\newline
\end{enumerate}

You can replace the station name "BONE" with any word of length less
than 32, and the network name "GR" with any word of length less than
8.
And you can always plot OUTPUT\_FILES/sr.vtk file in ParaView to check the
source/receiver locations after your simulation.\newline
\end{itemize}

@@

\begin{itemize}
\item The simulation is launched within the root directory \texttt{EXAMPLE\_ROOT\_DIR}\newline
(usually \texttt{mpirun -np N ./bin/xspecfem3D}).
\item \texttt{DATA} should contain the \texttt{Par\_file} parameter file with
\texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS} as explained in Chapter~\ref{cha:Creating-Distributed-Databases}.
\item \texttt{DATABASES\_MPI} and {OUTPUT\_FILES} directory may contain the mesher output but they
are not required as they are superseded by the ones in the \texttt{runXXX} directories.
\item \texttt{runXXXX} directories must be created beforehand. There should be be as many as
\texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS} and the numbering should be contiguous, starting from \texttt{0001}.
They all should have \texttt{DATA},
\texttt{DATABASES\_MPI} and \texttt{OUTPUT\_FILES} directories. Additionally a \texttt{SEM} directory
containing adjoint sources have to be created to perform adjoint simulations.
\item \texttt{runXXXX/DATA} directories must all contain a \texttt{CMTSOLUTION} file,
a \texttt{STATIONS} file along with an eventual \texttt{STATIONS\_ADJOINT} file.
\item If \texttt{BROADCAST\_SAME\_MESH\_AND\_MODEL} is set to \texttt{.true.} in \texttt{DATA/Par\_file},
only \texttt{run0001/OUTPUT\_FILES} and \texttt{run0001/DATABASES\_MPI} directories need to contain the files
outputted by the mesher.
\item If \texttt{BROADCAST\_SAME\_MESH\_AND\_MODEL} is set to \texttt{.false.} in \texttt{DATA/Par\_file},
every \texttt{runXXXX/OUTPUT\_FILES} and \texttt{runXXXX/DATABASES\_MPI} directories need to contain the files
outputted by the mesher. Note that while the meshes might have been created from different models and
parameter sets, they should have been created using the same number of MPI processes.
\end{itemize}

@@

\noindent
The model used is a constant $Q$, thus with no dependence on frequency ($Q(f)$ = constant).
See e.g. \cite{BlKoChLoXi16}. \newline

\noindent
However in practice for technical reasons it is approximated based on the sum of different Generalized Zener body mechanisms
and thus the code outputs the band in which the approximation is very good, outside of that range it can be less accurate.
The logarithmic center of that frequency band is the \texttt{ATTENUATION\_f0} parameter defined (in Hz) in input file \texttt{DATA/Par\_file}. \newline

### 2022-11-12 doc/USER_MANUAL/06_fault_sources.tex
updates tex files (\\ to \newline) for better pandoc conversions to wiki pages

Faults need to be handled in a special way during mesh generation.
A fault surface must lie at the interface between elements (the mesh
must honor the fault surfaces). Moreover, a fault is made of two surfaces
in contact. Each of these two surfaces needs a separate set of nodes.
This approach is known as \textquotedbl{}split nodes\textquotedbl{}.
Currently faults can only be run with \texttt{xdecompose\_mesh} and
\texttt{CUBIT}. \texttt{xmeshfem3D} is not yet ready to handle faults.\\

To facilitate the mesh generation with split nodes in CUBIT, we need
to separate the two fault surfaces by a small distance, effectively
creating a tiny opening of the fault (Figure \ref{fig:examples.splitnodes},
\ref{fig:examples.splitnodes-surfacetrace}). Note that the opening distance
should not be too small especially in single precision simulations
(For example, if your fault is at $y=y_0$ plane, then the opening distance
should be at least $10^{-6}y_0$ since single precision can ensure precision
within only 7 digits; furthermore, to properly recognize split nodes, the dimension
of the mesh also affect numerical roundoffs, and therefore the split nodes
should also be at least \texttt{10.d-10 * (Xmax-Xmin)} apart). Note that only the interior
of the fault must be opened, its edges must remain closed (except
the edge on the free surface). The fault is automatically closed later
by SPECFEM3D.\\

@@

\begin{figure}[htbp]
\begin{centering}
\includegraphics[scale=0.3]{figures/faultmesh.jpg} \\
\includegraphics[scale=0.3]{figures/surf168.jpg}
\includegraphics[scale=0.3]{figures/splitnodes.jpg}
\par
\end{centering}
\caption{Screenshots of the CUBIT example for the embedded fault with split
nodes: The entire mesh is decomposed into several volumes bounding
the fault (top), the fault surface 168 is shown as orange squares
(middle) and split nodes of the fault shown as orange dots(bottom).
Note that only interior nodes of the fault are split while those on
the edges of the fault surface are touching each other.}
\label{fig:examples.splitnodes}
\end{figure}

@@

\noindent
The CUBIT scripts ({*}.jou and {*}.py) in the directory \texttt{EXAMPLES}
generate more complicated meshes. The {*}.py files are Python scripts
that execute CUBIT commands and use the CUBIT-python interface for
SPECFEM3D (see next section). The Python language allows to define
and manipulate variables to parameterize the mesh. Alternatively,
the Python script can call a CUBIT journal file ({*}.jou), which looks
like the example above. Variables can be defined and manipulated there
using the $\mathtt{APREPRO}$ language built in CUBIT.\\

@@

The package includes examples, the SCEC benchmark problems:
\begin{itemize}
\item TPV5, a planar vertical strike-slip fault
\item TPV14 and TPV15, vertical strike-slip fault system with a fault branch
\item TPV102 and TPV103, rate-and-state benchmarks
\item TPV22 and TPV23, step-over benchmarks
\item TPV16, heterogenous initial stress
\end{itemize}
and
\begin{itemize}
\item Splay fault models from Wendt et al. (2009)\\

@@

During mesh generation, the fault is defined by two surfaces in contact.
Let's denote as \textquotedbl{}\emph{side 1}\textquotedbl{} the SECOND
surface declared by the user in the call to the python function \textquotedbl{}\emph{fault\_input}\textquotedbl{},
and the FIRST surface as \textquotedbl{}\emph{side 2}\textquotedbl{}.
The local coordinate system on the fault is defined as the right-handed
coordinate system defined by (strike, dip, normal), where \textquotedbl{}\emph{normal}\textquotedbl{}
is the normal vector outgoing from side 1, \textquotedbl{}\emph{dip}\textquotedbl{}
is parallel to the along-dip direction pointing downwards, and \textquotedbl{}\emph{strike}\textquotedbl{}
is the horizontal along-strike vector such that the system is right-handed. In places where the fault plane is horizontal,
we define the alone strike direction to be (1,0,0).\\

Slip is defined as displacement on side 2 minus displacement on side
1. In the local coordinate system on the fault, positive along-strike
slip is right-lateral and positive along-dip slip is thrust if side
1 is on the hanging wall (normal faulting if side 1 is on the foot
wall).\\

@@

Three additional input files are required in the \texttt{DATA/} directory
for dynamic and kinematic faults. They are $\mathtt{Par\_file\_faults}$,
$\mathtt{FAULT\_STATIONS}$ and $\mathtt{input\_file.txt}$ or $\mathtt{rsf\_hete\_input\_file.txt}$(optional).
If the former does not exist, the code assumes that there are no faults.\\

@@

The rest of this file is made of namelist input blocks (see \textquotedbl{}namelist\textquotedbl{}
in a Fortran 9x manual). The input for each fault has the following
sequence (arguments in {[}brackets{]} are optional):\\

\&\textbf{RUPTURE\_SWITCHES}  / RATE\_AND\_STATE ,TPV16 ,HETE\_RSF ,TPV10X ,TWF\\
\&\textbf{BEGIN\_FAULT} /\\
\&\textbf{STRESS\_TENSOR} Sigma = $\sigma_{xx}$,$\sigma_{yy}$,$\sigma_{zz}$,$\sigma_{xy}$,$\sigma_{xz}$,$\sigma_{yz}$ /\\
\&\textbf{INIT\_STRESS} S1, S2, S3 {[},n1, n2, n3{]} /\\

followed by (n1+n2+n3) \&\textbf{DIST2D} blocks\\

\&\textbf{SWF} mus, mud, dc {[}, nmus, nmud, ndc{]} (weakening\_kind=1 for linear (default); weakening\_kind=2 for exponential) /\\
\&\textbf{TWF} nuc\_x, nuc\_y, nuc\_z, nuc\_r, nuc\_t0, nuc\_v /\\
\&\textbf{RSF} V0,f0,a,b,L,V\_init,theta\_init,C,StateLaw {[} nV0,nf0,na,nb,nL,nV\_init,ntheta\_init,nC {]} /\\

followed by (nV0+nf0+na+nb+nL+nV\_init+ntheta\_init+nC) \&\textbf{DIST2D} blocks\\

@@

$\;$sets the name of the filename saved in the DATA directory.
The first line in the file shall be the number of data points and the
following lines in the file has the format:
x, y, z, val. Method of interpolation is nearest.
\\

@@

The purpose of the Kelvin-Voigt viscosity in the dynamic fault solver
is to damp spurious oscillations generated by the fault slip at frequencies
that are too high to be resolved by the mesh. The viscosity $\mathtt{eta}$
(in seconds) depends on the size of the elements on the fault. Here
is how to set it:
\begin{enumerate}
\item Determine the average linear size of the elements on the fault plane,
$\mathtt{h\_fault}$. Usually this value is prescribed by the user
during mesh generation. Otherwise it can be found by inspection of
the mesh inside the CUBIT GUI.
\item Use the Matlab function $\mathtt{utils/critical\_timestep.m}$ to
compute\\
 $\mathtt{dtc\_fault}=\mathtt{critical\_timestep\left(c_{p},h\_fault,ngll\right)}$.
\\
This is the critical time step in an elastic medium for a hypothetical
element of cubic shape with size equal to $\mathtt{h\_fault}$.
\item Set $\mathtt{eta}$ in $\mathtt{DATA/Par\_file\_faults}$ to (0.1 to 0.3)$\ensuremath{\times}\mathtt{dtc\_fault}$.
A larger $\mathtt{eta}$ damps high-frequencies more aggressively
but it might also affect lower frequencies and rupture speed.
\end{enumerate}
Viscosity reduces numerical stability: the critical timestep in a
simulation with Kelvin-Voigt damping needs to be smaller than that
in a purely elastic simulation. Here is how to set the time step accordingly:
\begin{enumerate}
\item Run a test simulation without viscosity ($\mathtt{eta}$=0 and only
a few time steps)
\item Look for the \textquotedbl{}maximum suggested time step\textquotedbl{}
in \texttt{OUTPUT\_FILES/output\_mesher.txt}. This is the critical
timestep of a purely elastic simulation, $\mathtt{dtc\_bulk}$.
\item Reset the timestep of the simulation with a Kelvin-Voigt material
to a value smaller than\\
 $\mathtt{dtc\_kv}=\mathtt{eta}\left(\sqrt{1+\mathtt{dtc\_bulk^{2}}/\mathtt{eta^{2}}}-1\right)$
\end{enumerate}
Note that in general $\mathtt{dtc\_bulk}$ is smaller than $\mathtt{dtc\_fault}$,
because elements off the fault might be smaller or more distorted
than element faces on the fault.

@@

Several output files are saved in \texttt{OUTPUT\_FILES/}:\\

\begin{enumerate}
\item Seismograms for each station on the fault plane given in \texttt{DATA/FAUL\_STATIONS}.
One output file is generated for each station, named after the station.
The files are ascii and start with a header (22 lines long) followed
by a data block with the following format, one line per time sample:\\

@@

Column 8 = normal stress (MPa)\\

@@

Some Matlab functions for post-processing and visualization are included
in directory $\mathtt{utils}$. The functions are internally documented
(see their matlab help).\\

\texttt{FSEM3D\_snapshot} reads a fault data snapshot\\

>>>>

Faults need to be handled in a special way during mesh generation.
A fault surface must lie at the interface between elements (the mesh
must honor the fault surfaces). Moreover, a fault is made of two surfaces
in contact. Each of these two surfaces needs a separate set of nodes.
This approach is known as \textquotedbl{}split nodes\textquotedbl{}.
Currently faults can only be run with \texttt{xdecompose\_mesh} and
\texttt{CUBIT}. \texttt{xmeshfem3D} is not yet ready to handle faults.\newline

To facilitate the mesh generation with split nodes in CUBIT, we need
to separate the two fault surfaces by a small distance, effectively
creating a tiny opening of the fault (Figure \ref{fig:examples.splitnodes},
\ref{fig:examples.splitnodes-surfacetrace}). Note that the opening distance
should not be too small especially in single precision simulations
(For example, if your fault is at $y=y_0$ plane, then the opening distance
should be at least $10^{-6}y_0$ since single precision can ensure precision
within only 7 digits; furthermore, to properly recognize split nodes, the dimension
of the mesh also affect numerical roundoffs, and therefore the split nodes
should also be at least \texttt{10.d-10 * (Xmax-Xmin)} apart). Note that only the interior
of the fault must be opened, its edges must remain closed (except
the edge on the free surface). The fault is automatically closed later
by SPECFEM3D.\newline

@@

\begin{figure}[htbp]
\begin{centering}
\includegraphics[scale=0.3]{figures/faultmesh.jpg} \newline
\includegraphics[scale=0.3]{figures/surf168.jpg}
\includegraphics[scale=0.3]{figures/splitnodes.jpg}
\par
\end{centering}
\caption{Screenshots of the CUBIT example for the embedded fault with split
nodes: The entire mesh is decomposed into several volumes bounding
the fault (top), the fault surface 168 is shown as orange squares
(middle) and split nodes of the fault shown as orange dots(bottom).
Note that only interior nodes of the fault are split while those on
the edges of the fault surface are touching each other.}
\label{fig:examples.splitnodes}
\end{figure}

@@

\noindent
The CUBIT scripts ({*}.jou and {*}.py) in the directory \texttt{EXAMPLES}
generate more complicated meshes. The {*}.py files are Python scripts
that execute CUBIT commands and use the CUBIT-python interface for
SPECFEM3D (see next section). The Python language allows to define
and manipulate variables to parameterize the mesh. Alternatively,
the Python script can call a CUBIT journal file ({*}.jou), which looks
like the example above. Variables can be defined and manipulated there
using the $\mathtt{APREPRO}$ language built in CUBIT.\newline

@@

The package includes examples, the SCEC benchmark problems:
\begin{itemize}
\item TPV5, a planar vertical strike-slip fault
\item TPV14 and TPV15, vertical strike-slip fault system with a fault branch
\item TPV102 and TPV103, rate-and-state benchmarks
\item TPV22 and TPV23, step-over benchmarks
\item TPV16, heterogenous initial stress
\end{itemize}
and
\begin{itemize}
\item Splay fault models from Wendt et al. (2009)\newline

@@

During mesh generation, the fault is defined by two surfaces in contact.
Let's denote as \textquotedbl{}\emph{side 1}\textquotedbl{} the SECOND
surface declared by the user in the call to the python function \textquotedbl{}\emph{fault\_input}\textquotedbl{},
and the FIRST surface as \textquotedbl{}\emph{side 2}\textquotedbl{}.
The local coordinate system on the fault is defined as the right-handed
coordinate system defined by (strike, dip, normal), where \textquotedbl{}\emph{normal}\textquotedbl{}
is the normal vector outgoing from side 1, \textquotedbl{}\emph{dip}\textquotedbl{}
is parallel to the along-dip direction pointing downwards, and \textquotedbl{}\emph{strike}\textquotedbl{}
is the horizontal along-strike vector such that the system is right-handed. In places where the fault plane is horizontal,
we define the alone strike direction to be (1,0,0).\newline

Slip is defined as displacement on side 2 minus displacement on side
1. In the local coordinate system on the fault, positive along-strike
slip is right-lateral and positive along-dip slip is thrust if side
1 is on the hanging wall (normal faulting if side 1 is on the foot
wall).\newline

@@

Three additional input files are required in the \texttt{DATA/} directory
for dynamic and kinematic faults. They are $\mathtt{Par\_file\_faults}$,
$\mathtt{FAULT\_STATIONS}$ and $\mathtt{input\_file.txt}$ or $\mathtt{rsf\_hete\_input\_file.txt}$(optional).
If the former does not exist, the code assumes that there are no faults.\newline

@@

The rest of this file is made of namelist input blocks (see \textquotedbl{}namelist\textquotedbl{}
in a Fortran 9x manual). The input for each fault has the following
sequence (arguments in {[}brackets{]} are optional):\newline

\&\textbf{RUPTURE\_SWITCHES}  / RATE\_AND\_STATE ,TPV16 ,HETE\_RSF ,TPV10X ,TWF\newline
\&\textbf{BEGIN\_FAULT} /\newline
\&\textbf{STRESS\_TENSOR} Sigma = $\sigma_{xx}$,$\sigma_{yy}$,$\sigma_{zz}$,$\sigma_{xy}$,$\sigma_{xz}$,$\sigma_{yz}$ /\newline
\&\textbf{INIT\_STRESS} S1, S2, S3 {[},n1, n2, n3{]} /\newline

followed by (n1+n2+n3) \&\textbf{DIST2D} blocks\newline

\&\textbf{SWF} mus, mud, dc {[}, nmus, nmud, ndc{]} (weakening\_kind=1 for linear (default); weakening\_kind=2 for exponential) /\newline
\&\textbf{TWF} nuc\_x, nuc\_y, nuc\_z, nuc\_r, nuc\_t0, nuc\_v /\newline
\&\textbf{RSF} V0,f0,a,b,L,V\_init,theta\_init,C,StateLaw {[} nV0,nf0,na,nb,nL,nV\_init,ntheta\_init,nC {]} /\newline

followed by (nV0+nf0+na+nb+nL+nV\_init+ntheta\_init+nC) \&\textbf{DIST2D} blocks\newline

@@

$\;$sets the name of the filename saved in the DATA directory.
The first line in the file shall be the number of data points and the
following lines in the file has the format:
x, y, z, val. Method of interpolation is nearest.
\newline

@@

The purpose of the Kelvin-Voigt viscosity in the dynamic fault solver
is to damp spurious oscillations generated by the fault slip at frequencies
that are too high to be resolved by the mesh. The viscosity $\mathtt{eta}$
(in seconds) depends on the size of the elements on the fault. Here
is how to set it:
\begin{enumerate}
\item Determine the average linear size of the elements on the fault plane,
$\mathtt{h\_fault}$. Usually this value is prescribed by the user
during mesh generation. Otherwise it can be found by inspection of
the mesh inside the CUBIT GUI.
\item Use the Matlab function $\mathtt{utils/critical\_timestep.m}$ to
compute\newline
 $\mathtt{dtc\_fault}=\mathtt{critical\_timestep\left(c_{p},h\_fault,ngll\right)}$.
\newline
This is the critical time step in an elastic medium for a hypothetical
element of cubic shape with size equal to $\mathtt{h\_fault}$.
\item Set $\mathtt{eta}$ in $\mathtt{DATA/Par\_file\_faults}$ to (0.1 to 0.3)$\ensuremath{\times}\mathtt{dtc\_fault}$.
A larger $\mathtt{eta}$ damps high-frequencies more aggressively
but it might also affect lower frequencies and rupture speed.
\end{enumerate}
Viscosity reduces numerical stability: the critical timestep in a
simulation with Kelvin-Voigt damping needs to be smaller than that
in a purely elastic simulation. Here is how to set the time step accordingly:
\begin{enumerate}
\item Run a test simulation without viscosity ($\mathtt{eta}$=0 and only
a few time steps)
\item Look for the \textquotedbl{}maximum suggested time step\textquotedbl{}
in \texttt{OUTPUT\_FILES/output\_mesher.txt}. This is the critical
timestep of a purely elastic simulation, $\mathtt{dtc\_bulk}$.
\item Reset the timestep of the simulation with a Kelvin-Voigt material
to a value smaller than\newline
 $\mathtt{dtc\_kv}=\mathtt{eta}\left(\sqrt{1+\mathtt{dtc\_bulk^{2}}/\mathtt{eta^{2}}}-1\right)$
\end{enumerate}
Note that in general $\mathtt{dtc\_bulk}$ is smaller than $\mathtt{dtc\_fault}$,
because elements off the fault might be smaller or more distorted
than element faces on the fault.

@@

Several output files are saved in \texttt{OUTPUT\_FILES/}:\newline

\begin{enumerate}
\item Seismograms for each station on the fault plane given in \texttt{DATA/FAUL\_STATIONS}.
One output file is generated for each station, named after the station.
The files are ascii and start with a header (22 lines long) followed
by a data block with the following format, one line per time sample:\newline

@@

Column 8 = normal stress (MPa)\newline

@@

Some Matlab functions for post-processing and visualization are included
in directory $\mathtt{utils}$. The functions are internally documented
(see their matlab help).\newline

\texttt{FSEM3D\_snapshot} reads a fault data snapshot\newline
